Reformulating phylogenetic mixed models to improve flexibility and speed

### 2018-08-29 main.tex
temp pushing

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
% Phylogenetic signal : the magnitude of phylogenetic correlation
The first challenge is segregating phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph? You go on to talk about the tip-variation/residual-variation confounding issue. Maybe replace with something talking more specifically about resolving tip variation?}
The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process. They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\ml{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}

The second challenge is flexibly model phylogenetic relationships to other predictor variables, that is, the variation among multiple different grouping predictor variables or variation in effects (random slopes) as well as correlated variation.
While it is reasonable to examine evolutionary relationships between species with their response, it is plausiable the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated. 
\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have tried to incorporate phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance. 
The tools available for extending phylogenetic relationships to predictor variables are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
Model & Method & Data & Platform \\
\hline
\hline
GLM (Simple)  & No Residual & Single Observation & GLS \\
\hline
              & Residual    & Single Observation & Pagel's $\lambda$ \\
              & + phylo intercept &                    & Blomberg's $K$ \\ 
              &             &                    & phylolm \\
\hline
GLMM (Complex)& Random effects & Single Observation & pez \\ 
              &                & Balanced Design &  \\
\hline
              &                & Unrestricted  & lme4, glmmTMB \\
\hline
Bayesian GLMM &                & Balanced Design & MCMCglmm \\ 
\hline        
              &                & Unrestricted   & brms \\
\hline
\end{tabularx}

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R package lme4 and glmmTMB with existing methods/platforms (i.e. nlme, phylolm, pez, MCMCglmm, and brms) to data from simulated model that incorportates phylogenetic signal from both predictors and tips/species, as well as residual variation.
In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

\newcommand{\bX}{{\boldmath X}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\bmu}{{\boldsymbol \mu}}
\newcommand{\by}{{\boldmath y}}
\newcommand{\bC}{{\boldmath C}}

The typical phylogenetic regression is of the form \bmb{boldface matrices/vectors i.e. $\bX \bbeta$? or not worth the trouble?
How about $\bmu=\bX \bbeta$; $\by \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality? (OK, I see that you move in this direction later.  But I'm not sure it's worth the detour of using two different notations? even if you are going to extend to include fixed effects, $Z$, etc. later ...}:
\begin{align}
y & = XB + \epsilon \label{eq:gls1} \\ 
\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C), \label{eq:gls2}
\end{align}
where $y$ is an $n \times 1$ response vector; $X$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $B$ is an $m$-vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ where is the $n \times n$ $C$ is the $n \times n$ phylogenetic correlation matrix.
The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
Y & \sim F(\mu) \label{eq:glmm1} \\
g(\mu) & = XB + Zb + \epsilon_{mm} \label{eq:glmm2} \\
\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2) \label{eq:glmm3}
\end{align}
where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$, $Z$ to an $n \times 1$ matrix such that:

\begin{align}
Z & \sim \textrm{MVN}(0,\sigma^{2}C). \label{eq:glmmgls}
\end{align}

\bmb{worth including the MCMCglmm/brms inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}

@@

An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$.

$S$ is a matrix product of an $n \times b$ indicator matrix $S_{id}$ of branch indices, and an adjusted branch length vector $B_{adj}$:

\begin{align}
B_{adj} & = (\frac{Bl}{\omega})^{1/2} \\
\omega & = \textrm{det}(\Sigma)^{1/n},
\end{align}
\bmb{would like to avoid multi-letter notation if possible; multi-letter subscripts should be text, e.g. ${\mathbf B}_{\mbox{\tiny adj}}$ (using macros to define stuff in one place \ldots) Also consider $\ell$ as a replacement for $l$}
where $Bl$ is the branch length vector, $\omega$ is a scaling factor, and $\Sigma$ is the variance-covariance matrix of the phylogenetic tree.
This is mathematically equivalent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.

The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \citet{bates2015fitting}.
Thus, the phylogenetic correlated random-effect matrix $Z^{C}_{i}$ is

\begin{equation}
Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}

where $\ast$ is the Khatri-Rao product \bmb{ref?}, $S_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.

@@

We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites/groups. 
To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$. \bmb{clarify what you mean by ``multiple levels of phylog. signals'' here?}
We further assume that random slope and intercept are correlated. \bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

@@

Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables. We implemented our approach using the \pkg{lme4} R package \citep{bates2015fitting}.
We compare our approach with three other R packages that can fit phylogenetic comparative models under the assumption that evolution follows BM process; are \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, and \pkg{pez} \citep{pearse2015pez}.
\bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
\begin{verbatim}
grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
invert=TRUE,value=TRUE)
\end{verbatim}
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
Phylogenetic generalized linear models (PGLM) (\bmb{function name?} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow for within-species variation (sometimes called the tip variation).
\bmb{does it handle non-Normal responses? I'd think so if it's ``generalized''. Worth establishing our terminology for ``tip variation''/''within-species variation'' etc. earlier?}
Neither PGLS nor PGLM can fit random-slope models or multiple sites \bmb{what does ``multiple sites'' mean in the model context?}
One of the relatively few packages that currently fit random slopes or multiple sites is \pkg{pez}, which can handle adding additional uncorrelated random slopes with intercept and species in multiple sites/groups.  
\bmb{I like the commented-out table here (maybe in supp material? or maybe useful enough that it's worth including in the main text?) although it could take a lot of fussing to get it right}

@@

% % \begin{table}
% % \begin{center}
% 
% \begin{tabularx}{\textwidth}{|X|X|X|}
% \hline
% Type of RE	& Formula	& Platform	\\
% \hline
% 
% Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, lme4 \\
% Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
% Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& lme4 \\
% \hline
% Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
% Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& pez, lme4 \\
% Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& lme4 \\
% \hline
% \end{tabularx}
% % \vspace{1in}
% % ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
% % \end{center}
% % \end{table}

@@

\bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order). Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods). Am I correct in believing that phylolm=lme4 but without intercept/slope correlation, and gls is additionally missing random slopes?}

@@

Platforms using the mixed model framework for multiple site simulations follow similar behaviour as the single-site simulation (i.e. different levels of phylogenetic signals and variations are closer to the simulated parameters with more data). 
All parameter estimates are much more robust for the multiple site than the single site simulations, especially slope-intercept correlations.
The lack of phylogenetic correlation in \pkg{pez}'s statistical models does not appear to hurt its ability to estimate other aspects of phylogenetic signal; in fact, its slightly simpler model may be responsible for its slightly less variable estimates.
\bmb{Can you identify the extreme cases in lme4's estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
However, despite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in \pkg{lme4} is dramatically better (orders of magnitude faster) than \pkg{pez}.
\bmb{be more specific, e.g. ``median of 8 vs 800 seconds for the medium-size example''. Clarify what ``independent errors'' means here?}
Furthermore, it was not practical for \pkg{pez} to fit large sample size simulation data as the computational speed did not scale linearly.
\bmb{Can you put a lower bound on pez's computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
  or (easier) do a brute-force series of increasing sample size and
  compute a log-log regression of time vs sample size to find the
  approximate power?}

@@

We used our method and fitted the full model of the dune meadow data recently used with pez \cite{li2017canfun}. 
We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to pez. 
Codes for fitting dune meadow analysis using lme4 and pez are provided in the supplements.

@@

Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis. 
This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as MCMCglmm \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e. What should we do when we don't have enough data? Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e. large data, unbalanced species observations, complex random effects) and explore new ideas.

>>>>

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
% Phylogenetic signal : the magnitude of phylogenetic correlation
The first challenge is segregating phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph? You go on to talk about the tip-variation/residual-variation confounding issue. Maybe replace with something talking more specifically about resolving tip variation?}
The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process. They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/interept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
\ml{make a diagram!}
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\ml{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}

The second challenge is flexibly model phylogenetic relationships to other predictor variables (predictor level variation), that is, the variation among multiple different grouping predictor variables or variation in effects (random slopes) (i.e phylogenetically related species may behave/ have similar changes in effect or in similar conditions) as well as correlated variation.
While it is reasonable to examine evolutionary relationships between species with their response, it is plausiable the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated. 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have tried to incorporate phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
Model & Method & Data & Platform \\
\hline
\hline
GLM (Simple)  & No Residual & Single Observation & GLS \\
\hline
              & Residual    & Single Observation & Pagel's $\lambda$ \\
              & + phylo intercept &                    & Blomberg's $K$ \\ 
              &             &                    & phylolm \\
\hline
GLMM (Complex)& Random effects & Single Observation & \pkg{pez} \\ 
              &                & Balanced Design &  \\
\hline
              &                & Unrestricted  & \pkg{lme4}, \pkg{glmmTMB} \\
\hline
Bayesian GLMM &                & Balanced Design & \pkg{MCMCglmm} \\ 
\hline        
              &                & Unrestricted   & \pkg{brms} \\
\hline
\end{tabularx}

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing methods/platforms (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorportates phylogenetic signal from both predictors and tips/species, as well as residual variation.
In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

\newcommand{\bX}{{\mathbf X}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\bmu}{{\boldsymbol \mu}}
\newcommand{\bY}{{\mathbf Y}}
\newcommand{\bC}{{\mathbf C}}
\newcommand{\bZ}{{\mathbf Z}}
\newcommand{\bb}{{\mathbf b}}
\newcommand{\bSigma}{{\boldsymbol \Sigma}}

The typical phylogenetic regression is of the form

%\bmb{boldface matrices/vectors i.e. $\bX \bbeta$? or not worth the trouble? How about $\bmu=\bX \bbeta$; $\bY \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality? (OK, I see that you move in this direction later.  But I'm not sure it's worth the detour of using two different notations? even if you are going to extend to include fixed effects, $Z$, etc. later ...}:

\begin{align}
\bmu & = \bX \bbeta  \label{eq:gls1} \\ 
\bY & \sim \textrm{MVN}(\bmu,\sigma^{2} \bC), \label{eq:gls2}
\end{align}

where $\bX$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector and assumed to be multivariate normally distributed with mean $\bmu$ and a variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods \citep{garamszegi2014modern}.

%\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
%\ml{chapter 7 in garamszegi 2014}

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
g(\bmu) & = \bX \bbeta + \bZ \bb \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3}
\end{align}
where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\bb$ is the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$.
Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$.

\bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
\ml{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}

@@

\newcommand{\bS}{{\mathbf S}}
\newcommand{\bJ}{{\mathbf J}}
\newcommand{\bBadj}{{\mathbf B}_{\mbox{\tiny adj}}}
\newcommand{\bomega}{{\boldsymbol \omega}}
\newcommand{\bell}{{\boldsymbol \ell}}

An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, rather than having to impose a correlation $\bC$.

$\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{id}$ of branch indices, and an adjusted branch length vector $\bBadj$:

\begin{align}
\bBadj & = (\frac{\bell}{\bomega})^{1/2} \\
\bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
\end{align}
% \bmb{would like to avoid multi-letter notation if possible; multi-letter subscripts should be text, e.g. ${\mathbf B}_{\mbox{\tiny adj}}$ (using macros to define stuff in one place \ldots) Also consider $\ell$ as a replacement for $l$}
where $\bell$ is the branch length vector, $\bomega$ is a scaling factor, and $\bSigma_{phy}$ is the variance-covariance matrix of the phylogenetic tree.
This is mathematically equivalent under BM assumption where we can reconstruct $\bC$ by matrix multiplying $\bS$ by itself.

The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrix $\bZ_{i}$ as described in \citet{bates2015fitting}.
Thus, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is

\begin{equation}
\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}
\ml{need a matrix diagram on the linear algebra}
where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is an $\ell_{i} \times b_{i}$ matrix of species--branch relations; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times \ell_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.

@@

We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species, and $m$ (1 and 25) sites/groups. 
To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$. \bmb{clarify what you mean by ``multiple levels of phylog. signals'' here?}
We further assume that random slope and intercept are correlated. \bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

@@

Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables. We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \ml{citation}.
We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}.
% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
% \begin{verbatim}
% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
% invert=TRUE,value=TRUE)
% \end{verbatim}
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow observational residuals.
Both packages can do other evolutionary models and different correlation structures (i.e. Pagel's $\lambda$, Blombergs $K$ and etc.)
Neither PGLS nor PGLM can multiple observations or predictor level variation \ml{intercept level vs predictor level?}. 
One of the relatively few packages that currently fit phylogenetic correlations to predictor level variationis is \pkg{pez}, which can handle adding additional uncorrelated random slopes and interactions with intercept.

\bmb{I like the commented-out table here (maybe in supp material? or maybe useful enough that it's worth including in the main text?) although it could take a lot of fussing to get it right}

@@

% % \begin{table}
% % \begin{center}
% 
% \begin{tabularx}{\textwidth}{|X|X|X|}
% \hline
% Type of RE	& Formula	& Platform	\\
% \hline
% 
% Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, \pkg{lme4} \\
% Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& \pkg{lme4} \\
% Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& \pkg{lme4} \\
% \hline
% Multiple site intercept 									&  1 $\mid$ sp 						& \pkg{pez}, \pkg{lme4} \\
% Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& \pkg{pez}, \pkg{lme4} \\
% Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& \pkg{lme4} \\
% \hline
% \end{tabularx}
% % \vspace{1in}
% % ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
% % \end{center}
% % \end{table}

@@

\bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order). Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods). Am I correct in believing that phylolm=\pkg{lme4} but without intercept/slope correlation, and gls is additionally missing random slopes?}

@@

Platforms using the mixed model framework for multiple site simulations follow similar behaviour as the single-site simulation (i.e. different levels of phylogenetic signals and variations are closer to the simulated parameters with more data). 
All parameter estimates are much more robust for the multiple site than the single site simulations, especially slope-intercept correlations.
The lack of phylogenetic correlation in \pkg{pez}'s statistical models does not appear to hurt its ability to estimate other aspects of phylogenetic signal; in fact, its slightly simpler model may be responsible for its slightly less variable estimates.
\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
However, despite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in \pkg{lme4} is dramatically better (orders of magnitude faster) than \pkg{pez}.
\bmb{be more specific, e.g. ``median of 8 vs 800 seconds for the medium-size example''. Clarify what ``independent errors'' means here?}
Furthermore, it was not practical for \pkg{pez} to fit large sample size simulation data as the computational speed did not scale linearly.
\bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
  or (easier) do a brute-force series of increasing sample size and
  compute a log-log regression of time vs sample size to find the
  approximate power?}

@@

We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to \pkg{pez}. 
Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.

@@

Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis. 
This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e. What should we do when we don't have enough data? Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e. large data, unbalanced species observations, complex random effects) and explore new ideas.
